\documentclass[fleqn,leqno]{article}
\usepackage{hypertlabook}
\pdftitle{Links}
\file{help-2}

\begin{document}
\noindent I don't like the use of the $\Box$, but I can't think of a
better way to give you the choice of opening a link in the same
window or in a new one.  It would be great if you could make the
choice by left- or right-clicking.  Perhaps someone will suggest that
to Adobe.  However, for now you can make the choice only by clicking
in one of two different places.

Clicking on the part of the link other than the $\Box$ might also
open a new window.  You can prevent that by going Adobe Reader's
\emph{Preferences} menu, reached from its main \emph{Edit} menu,
clicking on the \emph{Documents} category, and selecting the
\emph{Open cross-document links in same window} option in the
\emph{Open Settings} section.

If you opened this text in a new window, you can close that window the
same way you close other windows on your computer.  If you opened this
text in the same window as the link that brought you here, it would be
nice if you could return by clicking on the \backbutton\ button in the
left margin.  Unfortunately, that button may or may not work when
you jump to new text using a link like \rref{help-2}{nowhere}{this},
depending on your operating system, the version of Adobe Reader you
are using, and perhaps the weather in Katmandu.  It seems always to
work if you jump with a link like \lref{top}{this}.  To return to
where you came from, \sref{help}{help-2-return}{click here}.

% \medskip Note: the \backbutton\ button doesn't work the way it should
% because Adobe Reader's \emph{Preview View} command doesn't take you
% back to text in a different pdf file.  Someone might suggest to Adobe
% that this is a feature they might want to add.  (I have tried
% this only on Windows; 
%    \hyperref{http://lamport.org}{}{}{\linkcolor tell me}
% if it works on your operating system.)

\end{document}
